### 2016-09-06 doc/manuscript.Rmd
Edited methods section in manuscript

- Primary analysis:
    - Heatmap
    - GEE
        - Include # of both uncorrected and corrected (FDR) significant associations
        - QIC modeling
        - DAG development
        - Look over the entire model (see which variables are significant, etc)
    - PLS with ISI and HOMA-IR
        - CV for determining number of comps (best is 2)
        - Loadings (of TAGFA) (tho all are the same, use HOMA/ISI. BCF have
        lower corr between measured and predicted) (Use COrr of both components
        (1 and 2))
        - Prediction (of predicted vs real) (in the pls.rmd file, not in
        manuscript (or maybe in ESM?))
        - Use extracted PLS components in GEE (?)
        - Train and test (CV) to see how it predicts in a new dataset (ie. a
        future visit) (done. In pls.rmd file. As ESM?)
- Secondary analysis:
    - PLS-DA (dysgly)
        - Train and test (CV)
        - Loadings (of TAGFA) 
        - Groupings
        - Prediction (misclassification)
    - AROC
        - dysgly with clinical TAG vs the 'best' fatty acid to compare to

@@

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE required having one or more risk factors for type 2
diabetes mellitus, including obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=736) attended the baseline visit
between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013). 
Participants are contacted annually by telephone. The
current study sample used data on participants who did not have diabetes at baseline,
who returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477). A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`. At each examination,
participants undergo metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics. Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario,
and all participants provided written informed consent.
Data collection methods were standardized across the 2 centres and research
nurses were centrally trained.

## Blood measure assessments

At each examination, an 8-12 hour fasting blood sample was drawn from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and clinically-measured triacylglycerides (TAG)
were measured using Roche Modular's enzymatic colorimetric tests (Mississauga,
ON).  Both insulin and glucose were derived from the OGTT at fasting, 30 minute,
and 2 hour time points. Specific insulin was measured using the Elecsys 1010
(Roche Diagnostics, Basel, Switzerland) immunoassay analyzer and
electrochemiluminescence immunoassay. This assay shows 0.05% cross-reactivity to
intact human pro-insulin and the Des 31,32 circulating split form (Linco Res.
Inc), and has a coefficient of variation (CV) of 9.3%. Glucose was determined
using an enzymatic hexokinase (Roche Modular, Roche Diagnostics) with a
detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The inter-assay %CV is <1.1%
and intra-assay %CV is < 1.9%.  All assays were performed at the Banting and
Best Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose
(IFG), impaired glucose tolerance (IGT), and diabetes were categorized using the
2006 WHO criteria [@WHO2006].

TAGFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be 
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) was added to the serum as an internal standard
prior to extracting total lipids according to the method of Folch [@Folch1957a].
Each serum lipid fraction---non-esterified fatty acids (NEFA), cholesteryl
ester, phospholipid, and TAGFA---was isolated using thin layer chromatography;
each fraction was visualized under UV light after lightly spraying with
8-anilino-1-naphthalene sulfonic acid (0.1% wt/vol) and then converted to fatty
acid methyl esters with 14% boron trifluoride in methanol at 100°C for 1 h.
Fatty acid methyl esters were separated and quantified using a Varian-430 gas
chromatograph (Varian, Lake Forest, CA, USA) equipped with a Varian Factor Four
capillary column and a flame ionization detector, which were injected in
splitless mode. Fatty acid concentrations (nmol/ml) were calculated by
proportional comparison of gas chromatography peak areas to that of the internal
standards [@Nishi2014a]. There were 22 fatty acids measured in the TAGFA
fraction. Given their diverse biology, as well as the complexity of the
analyses, findings for other lipid fractions in this cohort are reported
separately (see ref @Johnston2016a and for the analysis of the phospholipid and
cholesteryl ester fractions and {{next citation}} for the analysis of the NEFA
fraction).

## Outcome variables

@@

Height, weight, and waist circumference (WC) were measured at all clinic 
examinations using standard procedures.  WC was measured at the natural waist, 
defined as the narrowest part of the torso between the umbilicus and the xiphoid
process. BMI was calculated by dividing weight (kg) by height (m) squared.
Sociodemographic information, including age, sex, and ethnicity, were determined
using questionnaires administered at each examination. In the lifestyle
questionnaire, physical activity was determined using a version of the
Modifiable Activity Questionnaire (MAQ) [@Kriska1990].  The MAQ collects 
information on leisure and occupational activity, including intensity, 
frequency, and duration, over the past year.  Each reported activity from the 
MAQ was weighted by its metabolic intensity allowing for the estimation of 
MET-hours per week.

## Statistical analysis

@@

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design 
an auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE models, 
though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
is well suited to longitudinal cohort studies given it's capacity to handle
missed visits. The predictor variables and continuous covariates were scaled
(mean centered and standardized). The TAGFA variables were classified as
*time-independent* (held constant) as they were measured only at the baseline
visit, while the outcome variables and covariates were set as *time-dependent*.
No imputation was conducted on missing values.

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC) {{citation?}}. `r cite_st('qic')` shows the covariates compared using QIC. While the
final GEE model selected as best fitting differed between insulin sensitivity
and beta-cell function measures, the differences in QIC values were less than
10 between the models, suggesting similar fit. As such, we selected the model
that had the fewest covariates and that had similar fit between the outcome
measures ({{##}}). The final GEE model was adjusted for {{covariates}}. After scaling,
log-transforming, and exponentiating, the GEE estimates are
interpreted as an expected percent difference in the outcome variable for every 
standard deviation (SD) increase in the predictor variable given the covariates 
are held constant. Lastly for the GEE models, we tested for an interaction with
sex, ethnicity, or time by the predictor term for each outcome variable.

@@

tagfa <- calc_tagfa_percent()
outcome_chg <- calc_outcome_changes()
calc_discr_npct(ds0yr$Sex, 'Female')
calc_discr_npct(ds0yr$Ethnicity, 'European')
estci <- calc_gee_estci(gee_df)
fa_cor <- calc_cor()
cor_pls_homa <- calc_pred_corr(pls_homa_t, pls_homa_t$test_data)
cor_pls_isi <- calc_pred_corr(pls_isi_t, pls_isi_t$test_data)
```

@@

@@

@@

>>>>

- Primary analysis:
    - Heatmap
    - GEE
        - Include # of both uncorrected and corrected (FDR) significant associations
        - QIC modeling
        - DAG development
        - Look over the entire model (see which variables are significant, etc)
        - Sensitivity analysis with Carb intake? And with adiponectin? (which is
        involved in TAG metabolism)
    - PLS with ISI and HOMA-IR
        - CV for determining number of comps (best is 2)
        - Loadings (of TAGFA) (tho all are the same, use HOMA/ISI. BCF have
        lower corr between measured and predicted) (Use COrr of both components
        (1 and 2))
        - Prediction (of predicted vs real) (in the pls.rmd file, not in
        manuscript (or maybe in ESM?))
        - Use extracted PLS components in GEE (?)
        - Train and test (CV) to see how it predicts in a new dataset (ie. a
        future visit) (done. In pls.rmd file. As ESM?)
- Secondary analysis:
    - PLS-DA (dysgly)
        - Train and test (CV)
        - Loadings (of TAGFA) 
        - Groupings
        - Prediction (misclassification)
    - AROC
        - dysgly with clinical TAG vs the 'best' fatty acid to compare to

@@

{{ Use an external source via DOI for reference to these methods? }}

Recruitment for the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
cohort baseline visit took place between 2004-2006 from London and Toronto, 
Canada. Individuals were selected to participant if they met the eligibility 
criteria of having one or more risk factors for type 2 diabetes mellitus, 
including obesity, hypertension, family history of diabetes, and/or a history of
gestational diabetes or birth of a macrosomic infant. A total of 736 individuals
attended the baseline visit. Subsequent examinations occurred every three years,
with three examination visits completed to date (2004-2006, 2007-2009, and
2010-2013). Participants are contacted annually by telephone. The current study
sample used data on participants who did not have diabetes at baseline, who
returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477; see the CONSORT diagram in `r
cite_sf('consort')`. Metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics are collected at each
examination visit. Research ethics approval was obtained from Mount Sinai
Hospital and the University of Western Ontario, and all participants provided
written informed consent. Data collection methods were standardized across the 2
centres and research nurses were centrally trained.

## Metabolic characterization

Participants completed an 8-12 hour 75g oral glucose tolerance test (OGTT) at 
each examination visit, with blood samples during at fasting, 30 min, and 2 hr 
post-glucose load. Samples were subsequently processed and frozen at -70°C.
Alanine aminotransferase (ALT) was measured using standard laboratory 
procedures. Cholesterol, HDL, and clinically-measured triacylglycerides (TAG) 
were measured using Roche Modular's enzymatic colorimetric tests (Mississauga, 
ON).  Both insulin and glucose were derived from the OGTT at fasting, 30 minute,
and 2 hour time points. Specific insulin was measured with the Elecsys 1010 
(Roche Diagnostics, Basel, Switzerland) immunoassay analyzer and 
electrochemiluminescence immunoassay, which shows 0.05% cross-reactivity to 
intact human pro-insulin and the Des 31,32 circulating split form (Linco Res. 
Inc) and has a coefficient of variation (CV) of 9.3%. Glucose was determined 
using an enzymatic hexokinase (Roche Modular, Roche Diagnostics) with a 
detection range of 0.11 to 41.6 mmol/L and an inter-assay %CV of <1.1% and an
intra-assay %CV of < 1.9%.  All assays were performed at the Banting and Best
Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose (IFG),
impaired glucose tolerance (IGT), and diabetes were categorized using the 2006
WHO criteria [@WHO2006].

TAGFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be 
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of {{ confirm with Richard }} heptadecanoic acid (17:0) was added as an internal
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction---non-esterified fatty acids (NEFA), 
cholesteryl ester, phospholipid, and TAGFA---was isolated using thin layer 
chromatography. Fatty acid methyl esters were separated and quantified using a 
Varian-430 gas chromatograph (Varian, Lake Forest, CA, USA) equipped with a 
Varian Factor Four capillary column and a flame ionization detector. Fatty acid 
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. There
were 22 fatty acids measured in the TAGFA fraction. Given that each lipid
fraction differs in biology and given the complexity of the analyses, findings
for other lipid fractions in this cohort are reported separately (see ref
@Johnston2016a and for the analysis of the phospholipid and cholesteryl ester
fractions and {{next citation}} for the analysis of the NEFA fraction).

Height, weight, and waist circumference (WC) were measured at all clinic 
examinations using standard procedures.  WC was measured at the natural waist, 
defined as the narrowest part of the torso between the umbilicus and the xiphoid
process. BMI was calculated by dividing weight (kg) by height (m) squared.
Questionnaires administered at each examination determined sociodemographics.
A version of the Modifiable Activity Questionnaire (MAQ) [@Kriska1990]
determined estimated physical activity. The MAQ collects information on leisure
and occupational activity, including intensity, frequency, and duration, over
the past year. Each reported activity from the MAQ was weighted by its
metabolic intensity allowing for the estimation of MET-hours per week.

## Variable calculation and statistical analysis

@@

@@

Generalized estimating equation (GEE) models [@Zeger1986a] were used in the
primary analysis to determine the longitudinal associations between the outcome
variables and the predictor variables. An auto-regressive of order 1 (AR1) 
correlation matrix was chosen for the GEE models as it had the best model fit
and given the longitudinal design, though other matrices (eg. exchangeable) had
similar fit (data not shown). GEE is well suited to longitudinal cohort studies
given it's capacity to handle missed visits. The predictor variables and
continuous covariates were scaled (mean centered and standardized). The TAGFA
variables were classified as *time-independent* (held constant) as they were
measured only at the baseline visit, while the outcome variables and covariates
were set as *time-dependent*. No imputation was conducted on missing values.

Covariates were selected based on previous studies, from directed acyclic graph 
[@Greenland1999a] recommendations, and from quasi-likelihood information criterion 
(QIC) {{citation?}}. `r cite_st('qic')` shows the covariates compared using QIC. While the
final GEE model selected as best fitting differed between insulin sensitivity
and beta-cell function measures, the differences in QIC values were less than
10 between the models, suggesting similar fit. As such, we selected the model
that had the fewest covariates and that had similar fit between the outcome
measures ({{##}}). The final GEE model was adjusted for {{covariates}}. After scaling,
log-transforming, and exponentiating, the GEE estimates are
interpreted as an expected percent difference in the outcome variable for every 
standard deviation (SD) increase in the predictor variable given the covariates 
are held constant. Lastly for the GEE models, we tested for an interaction with
sex, ethnicity, or time by the predictor term for each outcome variable.

@@

tagfa <- calc_tagfa_percent()
outcome_chg <- calc_outcome_changes()
female <- calc_discr_npct(ds0yr$Sex, 'Female')
euro <- calc_discr_npct(ds0yr$Ethnicity, 'European')
estci <- calc_gee_estci(gee_df)
fa_cor <- calc_cor()
cor_pls_homa <- calc_pred_corr(pls_homa_t, pls_homa_t$test_data)
cor_pls_isi <- calc_pred_corr(pls_isi_t, pls_isi_t$test_data)
```

@@

`r tab_basic` shows some basic character

@@

[@Lee2015a] 16:1n-7 may be good marker of liver DNL.

@@

```{r table_qic}
table_qic_main()
```
